Asymmetric functional prediction bands for assessing the validity of spatiotemporally smooth biomechanical curve data

### 2021-08-06 paper.Rmd
Comment out fda functions for the moment
Continue to work on the Rmd file (introduction section)

### TODO
- bar

@@

Knowing the size of the measurement error is a basic requirement for evaluating the results of empirical studies. Uncertainty about the error increases the risk of false positive (i.e. errors are incorrectly interpreted as effects) or false negative (i.e. effect and noise are inseperable) results and may negatively impact the reproducibility of study outcomes.

The classic linear error model describes measurement error as the sum of systematic and random deviations from the true value. The full characterization of measurement error, however, is considerably more complex, as different sources of variance (within, between) and noise must be considered. In particular, the output of many biomechanical measurement systems is available in the form of spatiotemporally smooth curves within which the error can vary over time.

However, to apply statistical standard models, in most cases the dimensionality of curves is reduced to just discrete features, such as local extrema, rates of change, or ranges of motion. A number of problems arise from this: (i) non-stationary error processes are not considered (ii) most traditional point statistics are either not suitable (e.g. statistical tests of mean differences), or are only suitable to a limited extent (e.g. product-moment correlation) to
quantify the agreement between two measurement systems (Bland and Altman, 1986) (iii) the validity of statements derived from discrete variables is limited by domain alignment (i. e. phase shift) and the risk of examining points that have little relevance for the system under investigation [Donoghue et al. (2008), Pataky et al. (2008), Richter et al. (2014), Park et al. (2017)].

Therefore, when analyzing the validity of continuous data, the entire length of the curve should be accounted for. This can be accomplished by either breaking down the curve into a single
value using e. g. distance measures such as the root mean square error, or by applying statistics (e. g. standard deviation) separately to all points of the curve [Pini et al. (2019)].

The pointwise approach allows for an error representation over the entire domain, but ignores the fact, that the sampling points of biomechanical curves are locally correlated [Deluzio et al. (2007), Pataky]. From a statistical point of view, this implies that random measurement error is likely underrepresented when the (nonzero) covariance term of adjacent sampling points is not accounted for [Lenhoff et al., 1999].

The principal object of this paper was to demonstrate a method for curve data that addresses the aforementioned issues (functional data representation, linear model assumptions, repeated measures), and thus enables the adequate characterization of the agreement between measurement curves on the basis of easy to interpret prediction bands (i.e. LoA in the actual measurement unit).

>>>>

### TODO
- Simulierte Daten durch Funktionen ersetzen? Eigentlich aktuell eher unnötig, da die Zeitreihen ja eigentlich bereits mithilfe von Funktionen modelliert wurden. Zudem wird nicht interpoliert, abgeleitet o. Ä. ... letztlich für die aktuelle Idee des Papers eher ungeeignet (plus man umgeht Probleme wie mögliche Glättungsartefakte, oder die Definition eines functional mean).
- Falls doch FDA, dann evtl. als eigenständiges Skript forken
- .bib file anlegen
- Begriff spatiotemporal überdenken (wie ist das im Kontext FDA definiert)
- Ratkowsky besorgen

### TODO read (ee 'Literature' folder on Nextcloud)
- Horvath, Kokoszka & Reeder (2013) Estimation of the mean of functional time series and a two-sample problem
- Generell die Kooperationspaper von Hörmann und Kokoszka zu FDA Statistik,insbesondere 'Inference for Functional Data with Applications' (2012)
- Ratkowsky 1990 Handbook of nonlinear regression models

@@

Having a good estimate of the amount of measurement error is a basic requirement when analyzing movement data. The classic linear error model describes measurement error as the sum of systematic and random deviations from the true value. A complete characterization of the measurement error, however, is complex, as different sources of variance (within, between) and noise must be considered. The analysis of sensor data is further complicated by the fact that the output of many measurement systems in biomechanics - such as e. g. joint angles, ground reaction forces, or center of mass trajectories - is available in the form of rather smooth curves within which the error can vary over time.

Therefore, when analyzing the validity of continuous data, the entire length of the curve should be accounted for. This can be accomplished e. g. by applying statistics (e. g. standard deviation) separately to all points of the curve [Pini et al. (2019)]. The pointwise approach allows for an error representation over the entire domain, but ignores the fact, that the sampling points of biomechanical curves are locally correlated [Deluzio & Astephen, 2007; Pataky, 2010]. From a statistical point of view, this implies that the random error component is likely underrepresented when the (nonzero) covariance term of adjacent sampling points is not accounted for [Lenhoff et al., 1999]. In addition, most traditional point statistics are either not suitable (e.g. statistical tests of mean differences), or are only suitable to a limited extent (e.g. product-moment correlation) to quantify the agreement between two measurement systems (Bland and Altman, 1986).

One possibility to address te problem of a pointwise analysis is to treat curves as single objects rather than as a series of unconnected points [Ramsay & Silverman, 2005]. Røislien et al. [2012] presented a method that extends the Limits of Agreement (LoA) concept [11] to functional data. This is achieved by approximating curves using Fourier series and resampling the thus obtained functions. The approach accounts for the autocorrelation characteristic in the signal and at the same time allows for an intuitive description of the systematic and random error terms in the original sampling space (e.g. actual degrees). Røislien et al. [10], however, make a number of strong assumptions, which are often not or only partially fulfilled in empirical data sets:

(i) Time-continuous differences of two measurement systems and the corresponding averages must be uncorrelated.
This assumption is violated considerably in many validation data sets [12]. E.g., Atallah et al. [13] found that the agreement between an ear-worn accelerometer and a force-plate instrumented treadmill was associated with gait speed when estimating gait cycle duration. When validating a foot-worn gyroscope for treadmill running, Koska & Maiwald [14] observed a linear error trend that was associated with foot strike behavior.

(ii) The error distribution is approximately normal.
The construction of symmetric prediction bands as described in Røislien et al. [10] relies on normally distributed measurement data. Non-normality, however, is often present when e.g. sensor values are truncated because of limited range or resolution, or because values drift over time.

(iii) Measurement uncertainty can be estimated adequately by bootstrapping the error variance from a sample that consists of a single curve per subject.
If no repeated measurements are included in the sample, an essential error component (intra-individual variance) is ignored. The bootstrap – like any other statistical procedure – cannot solve this problem, as it can only estimate proportions of variance that are contained in the sample.

Olsen et al. [15] examined the validity of an inertial measurement unit (IMU) for analyzing the spatial displacement of the distal limbs of horses using functional LoA (FLoA). Their approach differs from Røislien et al. [10] in that initially linear mixed models were fitted to the repeated measurements of different horses. Functional data estimates of the mean trends in the strides were then determined from the model fits and FLoA were calculated. This procedure accounts for repeated measurements and eliminates the need to emulate empirical curves with Fourier series or splines. The FLoA computed this way performed well in comparison with traditional Bland & Altman LoA. The procedure, however, is subject to the same restrictions as any conventional linear regression model and is therefore limited when model assumptions are violated.

The principal object of this paper was to demonstrate a method for curve data that addresses the aforementioned issues (analyzing the whole curve, linear model assumptions, repeated measures), and thus enables a more appropriate characterization of the agreement between measurement curves on the basis of easy to interpret prediction bands (i.e. LoA in the actual measurement unit).

We use synthetic time series data sets containing typical curves of biomechanical measurement systems to illustrate the method and compare the resulting prediction bands to pointwise Bland & Altman parameters.
